Longitudinal Manifold Estimation

### 2023-03-15 paper/main.tex
change otimes size

To ensure that the upcoming definitions are well behaved, we introduce the following function spaces:
\begin{itemize}
  \item $\mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D}) = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$
  \item $\nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d}) = \left\{f \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
  \item $H^2(\mathbb{R}^{d}) = \left\{f \in L^2(\mathbb{R}^{d}): \|\nabla f\|_{\mathbb{R}^{d}}, \|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
  \item $\nabla^{-\bigotimes 2}L^2(\Omega) = \left\{f|_{\Omega}: f \in \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d})\right\}$
  \item $H^2(\Omega) = \left\{f|_{\Omega}: f \in H^2(\mathbb{R}^{d})\right\}$
\end{itemize}

In these function spaces, we have the notation:
\begin{itemize}
  \item $\Omega$ is any open subset of $\mathbb{R}^{d}$ with a smooth boundary
  \item $f|_{\Omega}$ denotes the restriction of $f$ to $\Omega$
  \item $\nabla f$ denotes the gradient vector of $f$
  \item $\nabla^{\bigotimes 2}f$ denotes the Hessian matrix of $f$
  \item $\|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\bigotimes 2}f(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial t_i \partial t_j}(r)\right|^2\right)^{\frac{1}{2}}$
  \item $\|\nabla f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i=1}^{d}\left|\frac{\partial f}{\partial r_i}(r)\right|^2dr $
  \item $\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial r_i \partial r_j}(r)\right|^2dr $
\end{itemize}

To completely define principal manifolds, \cite{mengPrincipalManifoldEstimation2021} let $\|Pf\|_{\mathcal{H}} = \|f''(r)\|_{\mathbb{R}^{D}}^2$, which provides a greater penalization of roughness than using the first derivative. To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, they generalize this term as
\[%
  \|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr
,\]%
referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:
\begin{definition}
  \label{def:principal_manifolds}
  Let $X$ be a random $D$-vector associated with the probability measure or density function $\mathbb{P}$ such that $X$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments. Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
  \[%
    \mathcal{K}_{\lambda, \mathbb{P}}(f, g) = E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \quad \mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f)
  .\]%

  Then a manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ is called a principal manifold for $X$ with the tuning parameter $\lambda$ if 
  \[%
    f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \ \text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}
  .\]%
\end{definition}

@@

To estimate principal manifolds under this definition, \cite{mengPrincipalManifoldEstimation2021} propose the Principal Manifold Estimation (PME) algorithm. This algorithm has three main steps: data reduction, fitting, and tuning. In the reduction step, the HDMDE algorithm uses $k$-means clustering to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by cluster centers $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$. The number of clusters is chosen using an iterative hypothesis testing approach. For each given value of $\lambda$, the fitting stage of the algorithm uses this reduced dataset to find estimated function 
\[%
  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f)
,\]%
where 
\[%
  \mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2
\]% 
approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses the optimal value of $\lambda$, $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
\[%
  \mathcal{D}(\hat{f}_\lambda) = \frac{1}{I}\sum_{i=1}^{I}\|x_i - \hat{f}_\lambda(\pi_{\hat{f}_\lambda}(x_i))\|_{\mathbb{R}^{D}}^2
\]%
is an empirical estimate of the MSD using the full dataset.

>>>>

To ensure that the upcoming definitions are well behaved, we introduce the following function spaces:
\begin{itemize}
  \item $\mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D}) = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$
  \item $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{f \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
  \item $H^2(\mathbb{R}^{d}) = \left\{f \in L^2(\mathbb{R}^{d}): \|\nabla f\|_{\mathbb{R}^{d}}, \|\nabla^{\otimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
  \item $\nabla^{-\otimes 2}L^2(\Omega) = \left\{f|_{\Omega}: f \in \nabla^{-\otimes 2}L^2(\mathbb{R}^{d})\right\}$
  \item $H^2(\Omega) = \left\{f|_{\Omega}: f \in H^2(\mathbb{R}^{d})\right\}$
\end{itemize}

In these function spaces, we have the notation:
\begin{itemize}
  \item $\Omega$ is any open subset of $\mathbb{R}^{d}$ with a smooth boundary
  \item $f|_{\Omega}$ denotes the restriction of $f$ to $\Omega$
  \item $\nabla f$ denotes the gradient vector of $f$
  \item $\nabla^{\otimes 2}f$ denotes the Hessian matrix of $f$
  \item $\|\nabla^{\otimes 2}f\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\otimes 2}f(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial t_i \partial t_j}(r)\right|^2\right)^{\frac{1}{2}}$
  \item $\|\nabla f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i=1}^{d}\left|\frac{\partial f}{\partial r_i}(r)\right|^2dr $
  \item $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial r_i \partial r_j}(r)\right|^2dr $
\end{itemize}

To completely define principal manifolds, \cite{mengPrincipalManifoldEstimation2021} let $\|Pf\|_{\mathcal{H}} = \|f''(r)\|_{\mathbb{R}^{D}}^2$, which provides a greater penalization of roughness than using the first derivative. To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, they generalize this term as
\[%
  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr
,\]%
referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:
\begin{definition}
  \label{def:principal_manifolds}
  Let $X$ be a random $D$-vector associated with the probability measure or density function $\mathbb{P}$ such that $X$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments. Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
  \[%
    \mathcal{K}_{\lambda, \mathbb{P}}(f, g) = E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \quad \mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f)
  .\]%

  Then a manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ is called a principal manifold for $X$ with the tuning parameter $\lambda$ if 
  \[%
    f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \ \text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}
  .\]%
\end{definition}

@@

To estimate principal manifolds under this definition, \cite{mengPrincipalManifoldEstimation2021} propose the Principal Manifold Estimation (PME) algorithm. This algorithm has three main steps: data reduction, fitting, and tuning. In the reduction step, the HDMDE algorithm uses $k$-means clustering to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by cluster centers $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$. The number of clusters is chosen using an iterative hypothesis testing approach. For each given value of $\lambda$, the fitting stage of the algorithm uses this reduced dataset to find estimated function 
\[%
  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f)
,\]%
where 
\[%
  \mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2
\]% 
approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses the optimal value of $\lambda$, $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
\[%
  \mathcal{D}(\hat{f}_\lambda) = \frac{1}{I}\sum_{i=1}^{I}\|x_i - \hat{f}_\lambda(\pi_{\hat{f}_\lambda}(x_i))\|_{\mathbb{R}^{D}}^2
\]%
is an empirical estimate of the MSD using the full dataset.
